Visualizing demographic evolution using geographically inconsistent census data

### 2019-08-24 paper/related.tex
from overleaf

Since our problem encompasses several fields, we divide this section into
specific sub problems: \emph{longitudinal demographic studies}, describing the
traditional tabular approach to longitudinal studies; \emph{data
representation}, elaborating how evolving geographic data can be represented for
processing; \emph{data clustering and regionalisation}, briefly reviewing
existing data clustering methods,\revision{ geographic constraints and
regionalisation}; and \emph{cluster characterisation}, articulating how clusters
can be visually summarised.

\subsection{Longitudinal demographic studies}
Census data is used not only to discover demographic
patterns~\citep{Firebaugh2016}, but to correlate demographic characteristics to
other measurements~\citep{diez1997neighborhood}. However, longitudinal studies
are rare, because they are difficult : \emph{"(...) One of the most challenging
and fascinating areas in spatial statistics is the synthesis of spatial data
collected at different spatial scales(...)"}~\citep{gotway2002combining}. While
census tract level data is readily available for the US since at least
1910~\citep{nhgis}, most studies consider the period between 1970 and 2010,
using pre-harmonised data from the Longitudinal Tract Data
Base~\citep{Logan2014}. Despite its inherent
errors~\citep{Logan2016,Hallisey2017}, this dataset has become \revision{widely
adopted, along with the Neighborhood Change Database~\cite{ncdb},} as source for
longitudinal demographic data at the neighbourhood scale, with similar efforts
appearing in other countries~\citep{Liu2015,Lee2015,Allen2018}. These datasets
have been highly significant for the field. Yet they also limit the universe of
data that can be used to study neighbourhood change, since any new datasets
would need to be similarly processed in order to be rendered compatible with
these sources.

Another option considers the use of grid
data~\citep{Dmowska2017,Dmowska2018,stepinski2019imperfect}. Beyond the
potentially increased spatial precision, this approach does not require complex
harmonisation when new data is considered, if the grids are compatible. However,
demographic data is usually not available in this format, especially from older
sources. Additionally, the conversion from tabulation areas can introduce
significant errors.

@@

Data clustering is one of the elementary processes for data analysis,
simplifying the data into a smaller number of homogeneous sets that can be
interpreted in the same way. There is no shortage of contributions for this
problem~\citep{Fahad2014}\revision{, since variations of it appear in almost all
scientific fields.}

\revision{In geography, this problem is known as
\emph{regionalisation}~\citep{montello2003regions}, a rather old problem that
has been throughly explored, leveraging different mathematical tools, including
discrete topology~\citep{brantingham1978topological} and discrete
geometry~\citep{assunccao2006efficient}. Indeed, network-based methods are among
the current state-of-the-art~\citep{Guo2008Redcap,maxp}. However,
\emph{temporal} regionalisation is significantly less explored, especially in a
demographic context, arguably due to the difficulties in dealing with
unharmonised longitudinal data. Recent neighbourhood related applications rely
on k-means~\citep{jain2010data,Delmelle2016}, the Louvain method for community
detection~\citep{blondel2008Louvain,Thomas2012}, or, to a lesser extent, Self
Organising Maps~\citep{Delmelle2017,Ling2016,Dani2013_som_maxp}.

Since we adopted a network-based data representation and our objectives include
an interactive interface, we opted for an heuristic variation of the maximum
weighted matching algorithm called \emph{sorted maximal
matching}~\citep{markus2017}, because of its simplicity, customisability, and
fast computation times. This algorithm operates on weighted networks, where a
distance metric is between the nodes is associated with the edges. We adopted a
distance based on the data associated with each node. The algorithm merges
clusters based on these distances, creating an hierarchy instead of a fixed
number of clusters. This hierarchical result is rendered by the interface,
allowing the user to change the number of clusters without reprocessing.
Changing the clustering algorithm would lead to different results, but any
hierarchical network clustering method, and distance metric, can be used in our
framework. }

\subsection{Cluster characterisation}
While visualisation has gained prominence as a crucial component of scientific
discovery, justification, and communication~\cite{tufte1998visual}, visually
representing evolving spatial data is a challenging old
problem~\citep{monmonier1990strategies,andrienko2003exploratory,ferreira2015visual,Zheng2016}.

@@

Using data clustering, we can represent the region's cluster instead of all the
its variables~\citep{Alce2018,Valdivia2015,VonLandesberger2016}. While this
simplifies the geographic portion of the visualisation, it introduces the
problem of how to summarise the contents of each cluster. One traditional
approach is to use parallel coordinates plot~\citep{ferreira2015urbane}, but
these they can get cluttered representing similar clusters over several
variables. Further, for demographic applications, the clusters are usually
strongly characterised by a small subset of
values~\citep{Delmelle2016,Delmelle2017}. Therefore, in the proposed method, we
identify the variables that are most relevant to the characterisation of each
cluster. The distribution of values on that variable is then represented using a
boxplot, a well known statistical plot displaying basic properties of the
distributions.

>>>>

Since our problem encompasses several fields, we divide this section into
specific sub problems: \emph{longitudinal demographic studies}, describing the
traditional tabular approach to longitudinal studies; \emph{data
representation}, elaborating how evolving geographic data can be represented for
processing; \emph{data clustering and regionalisation}, briefly reviewing
existing data clustering methods,\revision{2.1}{ geographic constraints and
regionalisation}; and \emph{cluster characterisation}, articulating how clusters
can be visually summarised.

\subsection{Longitudinal demographic studies}
Census data is used not only to discover demographic
patterns~\citep{Firebaugh2016}, but to correlate demographic characteristics to
other measurements~\citep{diez1997neighborhood}. However, longitudinal studies
are rare, because they are difficult : \emph{"(...) One of the most challenging
and fascinating areas in spatial statistics is the synthesis of spatial data
collected at different spatial scales(...)"}~\citep{gotway2002combining}. While
census tract level data is readily available for the US since at least
1910~\citep{nhgis}, most studies consider the period between 1970 and 2010,
\revision{1.3}{using pre-harmonised data from the Longitudinal Tract Data
Base~\citep{Logan2014} or the Neighborhood Change Database~\citep{ncdb}, albeit
their inherent errors~\citep{Logan2016,Hallisey2017}.} Similar harmonisation
efforts appear in other countries~\citep{Liu2015,Lee2015,Allen2018}. These
datasets have been highly significant for the field. Yet they also limit the
universe of data that can be used to study neighbourhood change, since any new
datasets would need to be similarly processed in order to be rendered compatible
with these sources.

Another option considers the use of grid
data~\citep{Dmowska2017,Dmowska2018,stepinski2019imperfect}. Beyond the
potentially increased spatial\revision{1.4}{ precision}, this approach does not
require complex harmonisation when new data is considered, if the grids are
compatible. However, demographic data is usually not available in this format,
especially from older sources. Additionally, the conversion from tabulation
areas can introduce significant errors.

@@

Data clustering is one of the elementary processes for data analysis,
simplifying the data into a smaller number of homogeneous sets that can be
interpreted in the same way. There is no shortage of contributions for this
problem~\citep{Fahad2014}\revision{}{, since variations of it appear in almost all
scientific fields.}

\revision{2.1}{In geography, this problem is known as
\emph{regionalisation}~\citep{montello2003regions}, a rather old problem that
has been thoroughly explored, leveraging different mathematical tools, including
discrete topology~\citep{brantingham1978topological} and discrete
geometry~\citep{assunccao2006efficient}. Indeed, network-based methods are among
the current state-of-the-art~\citep{Guo2008Redcap,maxp}. However,
\emph{temporal} regionalisation is significantly less explored, especially in a
demographic context, arguably due to the difficulties in dealing with
unharmonised longitudinal data. Recent neighbourhood related applications rely
on k-means~\citep{jain2010data,Delmelle2016}, the Louvain method for community
detection~\citep{blondel2008Louvain,Thomas2012}, or, to a lesser extent, Self
Organising Maps~\citep{Delmelle2017,Ling2016,Dani2013_som_maxp}.}

\revision{1.5/2.2}{Since we adopted a network-based data representation and our objectives include
an interactive interface, we opted for an heuristic variation of the maximum
weighted matching algorithm called \emph{sorted maximal
matching}~\citep{markus2017}, because of its simplicity, customisability, and
fast computation times. This algorithm operates on weighted networks, where a
distance metric between the nodes is associated with the edges. We adopted a
distance based on the data associated with each node. The algorithm merges
clusters based on these distances, creating a hierarchy instead of a fixed
number of clusters. This hierarchical result is rendered by the interface,
allowing the user to change the number of clusters without reprocessing.
Changing the clustering algorithm would lead to different results, but any
hierarchical network clustering method, and distance metric, can be used in our
framework. }

\subsection{Cluster characterisation}
While visualisation has gained prominence as a crucial component of scientific
discovery, justification, and communication~\citep{tufte1998visual}, visually
representing evolving spatial data is a challenging old
problem~\citep{monmonier1990strategies,andrienko2003exploratory,ferreira2015visual,Zheng2016}.

@@

Using data clustering, we can represent the region's cluster instead of all the
its variables~\citep{Alce2018,Valdivia2015,VonLandesberger2016}. While this
simplifies the geographic portion of the visualisation, it introduces the
problem of how to summarise the contents of each cluster. One traditional
approach is to use parallel coordinates plot~\citep{ferreira2015urbane}, but
they can get cluttered when representing similar clusters over several
variables. Further, for demographic applications, the clusters are usually
strongly characterised by a small subset of
values~\citep{Delmelle2016,Delmelle2017}. Therefore, in the proposed method, we
identify the variables that are most relevant to the characterisation of each
cluster. The distribution of values on that variable is then represented using a
boxplot, a well known statistical plot displaying basic properties of the
distributions.

### 2019-08-24 paper/study.tex
from overleaf

In this section we present two illustrative scenarios, usingq census data from
the United States~\citep{nhgis} and
Canada\footnote{\url{http://datacentre.chass.utoronto.ca/census/}}, tabulated by
CTs,  from 1970 to 2010.

@@

\revision{Accidentaly, the US data for 2010 is actually not from the decennial
census, but from the ACS 2006-2010. Further, the regions selected do not
correspond to any pre-defined regions (metros, cities, census areas), but to
arbitrary regions defined around a location of interest. We selected a minimal
set of aspects for each country, and they are not similar to each other, out of
convenience, since the variable matching was a manual process. These caveats
would likely compromise any serious attempt on a comprehensive demographic
study, but this is not the objective of this experiment. These results are meant
to demonstrate the utility of the interface for understanding the evolutionary
dynamics of urban neighbourhoods. They also show the face validity of the
results generated by our novel network-based approach. Indeed, our results were
perfectly aligned with several other studies, despite these methodological
missteps, which could arguably be an indicator of its robustness.}

\subsection{Chicago}
Our first scenario examines Chicago, focusing on a region loosely following the
City's administrative borders. Its demographic composition is well explored in
the literature, with reports of racial divide and
gentrification~\citep{Delmelle2016,Delmelle2017,Hwang2014}, so we expect our
results to contain stable regions where the Race aspect is relevant, and some
degree of population change, with increasing income and education levels.

@@

The trajectories plot illustrates the process of gentrification, also
illustrated in Figure~\ref{fig:chiWorkflow}, progressively absorbing regions
from the orange cluster (White). This corroborates results from the literature
reporting that Black neighbourhoods are less likely to
gentrify~\citep{Hwang2014}. Moreover, this process appears to be unidirectional,
as indicated by the limited number of trajectories leaving the purple stream.
Next, we select the region that is gentrified in 2010, by clicking on the
corresponding rectangle in the trajectories plot, updating the information on
the maps and the details portion of the interface.

@@

We consider a region that corresponds approximately to the administrative border
of the current city of Toronto, using all seven available aspects with equal
weights. While Chicago was fairly stable, Toronto is known to be a more dynamic
and diverse city, with significant and increasing immigrant
population~\citep{hulchanski2007three,Fong2011}, especially
Asian~\citep{Fong2003}. Toronto is also known for a stable and well defined
Jewish community~\citep{Harold2018, Fong2011}. Therefore, we expect a
combination of stable and dynamic regions on the results, with Place of Birth,
Home Language, and Religion identified as relevant aspects. The results are
summarised in Figure~\ref{fig:to}, considering eight clusters.

@@

% \subsection{Los Angeles}

% We selected for this scenario a region around the metropolitan area of Los
% Angeles (LA), following urban density. While Chicago and Toronto demonstrated
% the abilities of our method to examine stable and dynamic cities, LA is considerably
% larger, both in population and area. Since our method identifies only the eight
% most distinct groups, we expect that some groups will not be identified,
% especially if their characterisation is similar to another groups. Further, our
% system does not include the census variables related to Hispanic heritage, which
% were only included in the newer censuses. However, from the literature, we
% expect to see some clusters where Race is an important aspect, including White,
% Black, and Asian~\citep{Reibel2003}.

% The summary of the results using all aspects with equal weights and eight
% clusters is illustrated in Figure~\ref{fig:la}. The full statistical description
% of the clusters is illustrated in Figure~\ref{fig:laAll}, where the most
% relevant aspect of each cluster is highlighted. From the trajectories plot, we
% can see that there is a large but shrinking cluster, depicted in green, one
% increasing cluster in purple, an almost constant orange cluster, a smaller but
% increasing pink cluster, and three other small clusters. The corresponding map
% illustrates where these clusters are located, and that they are somewhat
% geographically stable, with some movement between the green, orange, and purple
% clusters.

% \begin{figure}
%     \centering 
%     \includegraphics[width=0.975\linewidth]{LA.pdf}
%     \caption{Result for Los Angeles with 8 clusters, including three small and
%         ephemeral clusters. Cluster characterisation is displayed
%         in Figure~\ref{fig:laAll}.\label{fig:la}}
% \end{figure}

% From Figure~\ref{fig:laAll}, we can see that the green cluster is characterised
% by a high percentage of White population, low percentage of population in the
% lowest income bracket, mostly administrative occupations, and about 30\% of the
% population with four or more years of college. The orange cluster is
% characterised by a high percentage of Black population, with few people in the
% highest range of income and education. The purple cluster corresponds to a high
% concentration of "Other" in race, which includes Hispanic for this dataset, high
% concentration of Labourers, and low education and income. The pink cluster
% contains a high percentage of Asian population and about 30\% of the population
% with four or more years of college. The light green cluster contains very few
% people in the lower income bracket, mostly White population, with the highest
% percentage of population with four or more years of college, working
% administrative jobs, and a high concentration of singles. The yellow cluster
% represents Black population, with higher level of education and income, mostly
% working administrative jobs. The brown cluster represent a majority of single
% population, working administrative jobs with mostly low income. The dark grey
% cluster is characterised by all its population in the lowest income bracket, low
% education level, with a majority of White population. Since the extremes in the
% boxplots of the grey cluster are not significantly different, we can also
% surmise that this cluster is either small or homogeneous.

% \begin{figure*}
%     \centering 
%     \includegraphics[width=0.95\linewidth]{laAll.pdf}
%     \caption{Full characterisation of the eight clusters found for LA. The red
%      rectangles indicate the most relevant aspects for each
%      cluster.\label{fig:laAll}}
% \end{figure*}

% The  green, orange, and purple clusters present a significant intra-cluster
% variance in most variables, as indicated by extreme whiskers of the boxplots.
% While fifty percent of the CTs in the green cluster have between 20\% and 40\%
% of people in the lowest income bracket, that cluster also includes CTs where
% none and all the population belongs to that bracket. This might indicate that
% this cluster represents different groups of people that are not different enough
% to be separated at this level of the hierarchy. Conversely, the light green,
% brown, and dark grey clusters are different enough to be separated into their
% own clusters at this level, despite being small and ephemeral, including only a
% few CTs.

% The orange area in the map in Figure~\ref{fig:la} presents movement, indicated by
% the presence of green and purple tones mixed with the orange, which may warrant
% further exploration. By clicking on the orange bar above the boxplot, we select
% all trajectories that contain the orange cluster. The corresponding details are
% illustrated in Figure~\ref{fig:labh}. This shows a location change, where the
% orange cluster is progressively replaced by the purple cluster on its east side,
% and in turn expanding to the west. Interestingly, the population increased, the
% racial profile changed, but the distribution of income was reasonably stable,
% with a higher amount of the population in the lowest income range and very few
% people in the highest income range. Indeed, the income  difference is
% significant when compared to the city-wide distribution.

% \begin{figure}
%     \centering 
%     \includegraphics[width=0.75\linewidth]{laBH.pdf}
%     \caption{Top: Geographic changes in the majority Black population cluster
%     (orange) and Labourers cluster (green). Bottom: Income evolution for this
%     region (black) and the whole city (grey).\label{fig:labh}}
% \end{figure}

% A portion of this region is classified into the green cluster in 2010,
% indicating a majority white population. To further understand that change, we
% clear the current selection, and select all regions that changed from orange in
% 1970 to green in 2010, using the transition matrix. A portion of the resulting
% region, near the Florence-Graham region, is depicted in Figure~\ref{fig:laVol},
% along with the temporal evolution of Race. Despite this difference, the other
% aspects are similar to the ones from the region in Figure~\ref{fig:labh}, with
% slightly lower income and education profiles. While the racial aspect changed
% considerably, the economic and educational aspects stayed the same.

% \begin{figure}
%     \centering 
%     \includegraphics[width=0.75\linewidth]{laVol.pdf}
%     \caption{Details for a volatile region contained in the area of
%          Figure~\ref{fig:labh}. This region went from Black to Hispanic to
%         White.\label{fig:laVol}}
% \end{figure}

% These results are visually similar to the analysis presented by
% Dellmelle~\citep{Delmelle2016}.  What was there identified as 'persistently
% struggling' corresponds to the purple cluster in Figure~\ref{fig:la}; some
% portions of the 'stable elite' cluster, near Inglewood, correspond to the high
% income, high education yellow cluster. The other groups identified by that study
% were clumped by our system into the green cluster, which would be further
% divided if more than eight clusters were to be used.

% If Toronto illustrates a relatively smooth process of whole-sale
% diversification, Los Angeles shows the ability of our system and method to
% uncover more intensive smaller-scale processes of neighbourhood change. This
% capacity to examine and classify different types of change operating at
% different scales demonstrates the flexibility of the method and tool for
% advancing the more open-ended notion of "neighbourhood" envisioned by recent
% neighbourhood research.

% \smallskip While Toronto is more dynamic than Los Angeles, possibly due to size
% differences, the volatile regions shown in Figure~\ref{fig:toVol} did not change
% as quickly or dramatically as the ones shown in Figure~\ref{fig:laVol}, which
% involved twice as many people. We found this trend to be related to the
% countries themselves, Canadian cities have larger areas undergoing slow, gradual
% changes, whereas American cities have more general stability, but quicker
% changes in smaller scales. The supplementary material contains brief summaries
% of all the regions accessible in this prototype.

@@

The experts also mentioned the poor responsiveness of the method when changes in
the clustering parameters required server-side processing (B,D). Indeed, the
current implementation can take a few minutes to cluster regions with high
number of CTs, like Los Angeles or Brooklyn. Server-processing reduced the
amount of data transferred to client, but it might increase the response time
under load. We implemented a caching policy that greatly improved the
performance, but fully pre-processing the results is not practical due to size
of the parameter space.

>>>>

In this section we present two illustrative scenarios, using census data from
the United States~\citep{nhgis} and
Canada\footnote{\url{http://datacentre.chass.utoronto.ca/census/}}, tabulated by
CTs,  from 1970 to 2010.

@@

\unsure{CHECK}\revision{1.12.(1-5)}{These results are meant to demonstrate the
utility of the interface for understanding the evolutionary dynamics of urban
neighbourhoods. They also show the face validity of the results generated by our
novel network-based approach. However, the US data for 2010 is  not from the
decennial census, but from the ACS 2006-2010, compromising the temporal
stability of the data. Further, the regions selected do not correspond to any
pre-defined regions (metros, cities, census areas), but to somewhat arbitrary
regions defined around a location of interest, making direct comparisons harder.
Since  variable matching is a  manual process, we selected a minimal set of
aspects for each country, which are not necessarily similar to each other, or
similar to studies in the current literature. These factors would seriously
hinder any study using the existing methods, but our results were well-aligned
with several other studies, illustrating the robustness of our approach to
imperfect conditions. }

\subsection{Chicago}
Our first scenario examines Chicago, using an arbitrary region larger than the
administrative borders. Its demographic composition is well explored in the
literature, with reports of racial divide and
gentrification~\citep{Delmelle2016,Delmelle2017,Hwang2014}, so we expect our
results to contain stable regions where the Race aspect is relevant, and some
degree of population change, with increasing income and education levels.

@@

The trajectories plot illustrates the process of gentrification, also
illustrated in Figure~\ref{fig:chiWorkflow}, progressively absorbing regions
from the orange cluster (White). This corroborates results from the literature
reporting that Black neighbourhoods are less likely to
gentrify~\citep{Hwang2014}. The transition matrix provides more details, showing
that around one percent of all regions that became purple at any time
transitioned from the green cluster (Black population), while nine percent were
previouly associated with the orange cluster (White). Next, we select the region
that is gentrified in 2010, by clicking on the corresponding rectangle in the
trajectories plot, updating the information on the maps and the details portion
of the interface.

@@

We consider a region that corresponds approximately to the administrative border
of the current city of Toronto, using all seven available aspects with equal
weights. While Chicago was fairly stable, Toronto is known to be a more dynamic
and diverse city, with significant and increasing immigrant
population~\citep{hulchanski2007three,Fong2011}, especially
Asian~\citep{Fong2003}. Toronto is also known for a stable and well defined
Jewish community~\citep{Harold2018, Fong2011}. Therefore, we expect \revision{1.14}{a}
combination of stable and dynamic regions on the results, with Place of Birth,
Home Language, and Religion identified as relevant aspects. The results are
summarised in Figure~\ref{fig:to}, considering eight clusters.

@@

@@

The experts also mentioned the poor responsiveness of the method when changes in
the clustering parameters required server-side processing (B,D). Indeed, the
current implementation can take a few minutes to cluster regions with high
number of CTs, like Los Angeles or Brooklyn. Server-side processing reduced the
amount of data transferred to client, but it might increase the response time
under load. We implemented a caching policy, but fully pre-processing the
results is not practical due to size of the parameter space.
